Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-07 talk/dls2012/paper.tex
I'm *pretty* sure that this needs to be J, not I
some more typos, and three references

If a pure operation appears more than once in the trace with the same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. RPython's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{XXX}. In addition writes on newly allocated objects
can never change the value of old existing ones.}

@@

Consider the \lstinline{get} operation on line 22 of
Figure~\ref{fig:peeled-trace}. The result of this operation can be
deduced to be $i_3$ from the \lstinline{get} operation on line
8. The optimization will thus remove line 22 from the trace and
replace $i_7$ with $i_3$. Afterwards the trace is no longer in the correct
form, because the argument $i_3$ is not passed along the loop arguments. It
thus needs to be added there.

The trace from Figure~\ref{fig:peeled-trace} will therefore be optimized to:

@@

In general what is needed is to keep track of
which variables from the preamble are reused in the peeled loop.
By constructing a vector, $H$,  of such variables, the input and jump
arguments can be updated using
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H|}\right)
  \label{eq:heap-inputargs}
\end{equation}
and
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H|})\right)
  .
  \label{eq:heap-jumpargs}
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$.

@@

RPython's allocation removal optimization~\cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. That is, no outside
object ever gets a reference to them. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes.\footnote{This is sometimes called \emph{scalar replacement}.}
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed from $\hat J$ using the map $m$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

The current approach still has some limitations which we plan to address in the
future. In particular loop peeling works poorly in combination with trace
trees or trace stitching. The side exits attached guards that fail often
currently have to jump to the preamble which makes loops with several equally
common paths less efficient than they could be.

>>>>

If a pure operation appears more than once in the trace with the same input
arguments, it only needs to be executed the first time and then the result
can be reused for all other appearances. RPython's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{XXX}. In addition writes on newly allocated objects
can never change the value of old existing ones.}

@@

Consider the \lstinline{get} operation on line 22 of
Figure~\ref{fig:peeled-trace}. The result of this operation can be
deduced to be $i_3$ from the \lstinline{get} operation on line
8. The optimization will thus remove line 22 from the trace and
replace $i_7$ with $i_3$. Afterwards the trace is no longer in the correct
form, because the argument $i_3$ is not passed along the loop arguments.
Therefore $i_3$ needs to be added to the loop arguments.

Doing this, the trace from Figure~\ref{fig:peeled-trace} will be optimized to:

@@

In general what is needed is to keep track of
which variables from the preamble are reused in the peeled loop.
By constructing a vector, $H$,  of such variables, the input and jump
arguments can be updated using
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H|}\right)
  \label{eq:heap-inputargs}
\end{equation}
and
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H|})\right)
  .
  \label{eq:heap-jumpargs}
\end{equation}
In the optimized trace $J$ is replaced by $\hat J$ and $K$ by $\hat
K$.

@@

RPython's allocation removal optimization~\cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. That is, no outside
object ever gets a reference to them. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard_class}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes.\footnote{This is sometimes called \emph{scalar replacement}~\cite{kotzmann_escape_2005}.}
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed from $\hat J$ using the map $m$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

The current approach still has some limitations which we plan to address in the
future. In particular loop peeling works poorly in combination with trace
trees~\cite{andreas_gal_incremental_2006} or trace stitching~\cite{gal_trace-based_2009}.
The side exits attached guards that fail often
currently have to jump to the preamble which makes loops with several equally
common paths less efficient than they could be.
